---
title: "ml_epl"
author: "Alex Foster, Chris Bajek, Webster An"
date: "4/23/2020"
output: html_document
---

```{r setup, include=FALSE}

# Load packages
library(ggplot2)
library(dplyr)
library(janitor)
library(reshape2)
library(tidyr)
library(readr)
library(lubridate)
library(Matrix)

```
##### Analyze English Premier League Data with Machine Learning
#### Research Question: Can we predict the outcome (Home win, Away win or Draw) of the a Premier League match. 
* We intend on using multiple seasons. Our data source (https://datahub.io/sports-data/english-premier-league#data) has data for the past 10 years. 
* Below we have begun adding explanatory variables and created visualizations for two seasons of matches.


#### Load in the data
```{R}
epl_1819 <- read_csv("season-1819_csv.csv") %>%
  mutate(season = 1819, Datetime = dmy(Date),HAGG = 0.0, AAGG=0.0, GP = 0, HTP = 0, ATP=0, HACG=0.0,AACG=0.0) %>%
  select(Datetime, HomeTeam, AwayTeam, FTR,FTAG, FTHG, HTAG, HTHG, HTR,season, HAGG, AAGG,GP, ATP, HTP, HACG, AACG)
epl_1718 <- read_csv("season-1718_csv.csv") %>%
  mutate(season = 1718, Datetime = dmy(Date),HAGG = 0.0, AAGG=0.0, GP = 0, HTP = 0, ATP=0, HACG=0.0,AACG=0.0) %>%
  select(Datetime, HomeTeam, AwayTeam, FTR,FTAG, FTHG, HTAG, HTHG, HTR,season, HAGG, AAGG,GP, ATP, HTP, HACG, AACG)
epl_1617 <- read_csv("season-1617_csv.csv") %>%
  mutate(season = 1617, Datetime = dmy(Date),HAGG = 0.0, AAGG=0.0, GP = 0, HTP = 0, ATP=0, HACG=0.0,AACG=0.0)%>%
  select(Datetime, HomeTeam, AwayTeam, FTR,FTAG, FTHG, HTAG, HTHG, HTR,season, HAGG, AAGG,GP, ATP, HTP, HACG, AACG)

seasons = rbind(epl_1617,epl_1718)
seasons = rbind(seasons, epl_1819)
print(unique(seasons$season))
head(seasons)
```

#### Data Wrangling
* Need to make a dataframe where each row represents a game and has the following columns 
  * HT -- Home Team
  * AT -- Away Team
  * HTR -- Half Time Result
  * FTR -- Full Time Result
  * HTHG -- Half Time Home Goals
  * HTAG -- Half Time Away Goals
  * HAGG --  Home Average Goals per Game
  * AAGG -- Away Average Goals per Game 
  * HACG -- Home Average Conceded per Game
  * AACG -- Away Average Conceded per Game
  * HTP -- Home Team Position in the league
  * ATP -- Away Team Position in the league
  * Datetime -- lubricate datetime object 
* Response variable: FTR
* Explanatory variables: HTR, HTHG, HTAG, HAGG, AAGG, HTP, ATP 


```{r}
seasons_df <- seasons
game = seasons_df[40,]
print(game)
current_date <- game$Datetime
df_subset <- df[which(df$season == year(current_date) & df$Datetime <= current_date),]
df_subset
```

```{r}




# Code below computes the table of the season given a certain date in the season.
# Returns a dataframe that provides stats dependent on the progress of the given season 
# Tried to write a function but it wouldn't knit.
compute_table <- function(df = seasons_df, date, season) {
    df_subset <- df[which(df$Datetime <= date),] # get games from that season and up to the given data
    names <- unique(df_subset$HomeTeam) # get team names
    
    # initialize beginning of season table
    table <- data.frame("Team" = names, "Points" = matrix(0,length(names)), "GA" = matrix(0,length(names)), "GF" = matrix(0,length(names)), 'GP' = matrix(0,length(names)))
    
    for (row in 1:nrow(df_subset)) {
      current_r =df_subset[row, ]
      # Update info for home team
      home <- current_r$HomeTeam
      table$GA[which(table$Team == home)] <- table$GA[which(table$Team == home)] + current_r$FTAG# GA
      table$GF[which(table$Team == home)] <- table$GF[which(table$Team == home)]+current_r$FTHG# GF
      table$GP[which(table$Team == home)] <- table$GP[which(table$Team == home)]+1
    
      # Update info for away team
      away <- current_r$AwayTeam
      table$GA[which(table$Team == away)] <- table$GA[which(table$Team == away)]+current_r$FTHG# GA
      table$GF[which(table$Team == away)] <- table$GF[which(table$Team == away)]+current_r$FTAG# GF
      table$GP[which(table$Team == away)] <- table$GP[which(table$Team == away)]+1
    
      # Award points to winner and for ties.
      if (current_r$FTR == 'H'){ # home wins
        table$Points[which(table$Team == home)] <- table$Points[which(table$Team == home)] + 3
      } else if (current_r$FTR == 'A'){ # away wins
        table$Points[which(table$Team == away)] <- table$Points[which(table$Team == away)] + 3
      } else { # draw
        table$Points[which(table$Team == home)] <- table$Points[which(table$Team == home)] + 1
        table$Points[which(table$Team == away)] <- table$Points[which(table$Team == away)] + 1
      }
    }
    # compute goal differential
    table <- table %>%
      mutate('GD' = GF - GA)
    # add AGSG column that represents the avg goals scored per game
    table <- table %>%
      mutate(AGSG = GF / GP)
    # add AGCG column that represents the avg goals conceded per game
    table <- table %>%
      mutate(AGCG = GA / GP)
    # Add position column that indicates what place each team is in.
    table <- table[with(table,order(table$Points, table$GD)),]
    table$Position <- length(names):1
  
    return(table)
}
```

```{r}
# testing compute_table function
table = compute_table(seasons_df,current_date, 1617)  
head(table)
```


```{r}
# Function takes in dataframe with all matches of a SINGLE season. Then loops through each game For every match it computes the league standings prior to the given game
# Uses league standings to add more explanatory variables (HAGG and AAGG).
# Returns a dataframe of the games with each game updated for it's season statistics.
add_explantory_var <- function(df) {
    for (row in 1:nrow(df)){ # iterate through every game
        cur_season <- unique(df$season) # for computing the table

        current_date = df[row,]$Datetime # current date is the date of the game 
        
        if (month(current_date) == 8){
          next
        }
        # compute table
        table <- compute_table(df, current_date, cur_season)

        # Check if only 1 game has been played. Not worth computing stats if it's the first game of the season.
        if (table[1,]$GP != 1) {
            game = df[row,]
            # update HAGG, HACG, AACG and AAGG values for each game dateframe
            if (game$HomeTeam %in% table$Team & game$AwayTeam %in% table$Team) { # checks that both teams are in the table results.
                # home team stats
                df[row,]$HAGG <- table$AGSG[which(table$Team == game$HomeTeam)]
                df[row,]$HACG <- table$AGCG[which(table$Team == game$HomeTeam)]
                df[row,]$HTP <- table$Position[which(table$Team == game$HomeTeam)]
                # away teams stats
                df[row,]$AAGG <- table$AGSG[which(table$Team == game$AwayTeam)]
                df[row,]$AACG <- table$AGCG[which(table$Team == game$AwayTeam)]
                df[row,]$ATP <- table$Position[which(table$Team == game$AwayTeam)]

            }
        }
  }
  return(df)
}



```


```{r}
df_1617 <- add_explantory_var(seasons_df[which(seasons_df$season == 1617),])
df_1718 <- add_explantory_var(seasons_df[which(seasons_df$season == 1718),])
df_1819 <- add_explantory_var(seasons_df[which(seasons_df$season == 1819),])
final_df <- rbind(df_1617, df_1718)
final_df <- rbind(final_df, df_1819)
final_df

```

##### Putting it all together
```{r}
initialize_season <- sort(unique(seasons_df$season))[1]
final_df <- add_explantory_var(seasons_df[which(seasons_df$season == initialize_season),]) # initialize with the teams home games from the 1617 season
# loop through each season in the dataset.
for (season in unique(seasons_df$season)) {
    if (season != initialize_season) {
        df = seasons_df[which(seasons_df$season == season),]
        update_df = add_explantory_var(df)

       final_df = rbind(final_df, update_df)
    }
}



cat('Final dataset includes: ')
cat("\n")
cat("Seasons:",unique(final_df$season))
cat("\n")
cat("Rows:",nrow(final_df))

```

#### Data Visualizations

```{r}
# Distrbution of full time results by season
final_df %>% 
  ggplot(aes(x = FTR)) +
  geom_bar(colour = "#0c4c8a") +
  facet_wrap(~season)+
  theme_minimal()

# Distribution of full time results based on the HTR by season
final_df %>%
  ggplot(aes(x = FTR, fill = HTR)) +
  geom_bar() +
  scale_fill_hue() +
  facet_wrap(~season)+
  theme_minimal()

# Examing goals scored by away and home teams by date and facet by season  
ggplot(final_df) +
 aes(x = Datetime, y = FTHG) +
 geom_line(size = 1L, colour = "#0c4c8a") +
 theme_minimal() +
 facet_wrap(~season)
ggplot(final_df) +
 aes(x = Datetime, y = FTAG) +
 geom_line(size = 1L, colour = "#0c4c8a") +
 theme_minimal() +
 facet_wrap(vars(season))

# Examine average goals scored as the season progresses
# final_df %>%
#   ggplot(aes(x=Datetime, y=HAGG)) +
#   geom_line() +
#   facet_wrap(~season)
# final_df %>%
#   ggplot(aes(x=Datetime, y=AAGG)) +
#   geom_line() +
#   facet_wrap(~season)

```

#### Modeling 
* Plan is to focus on one team. 
  * Everton. 

Predict FTR variable which is either H (home win), A (away win) or D (draw)

##### Create training/testing set
* Going to have most recent season be testing set.
```{r}
df_modeling <- final_df %>%
  select(HomeTeam, AwayTeam, FTR, HTAG, HTHG, HTR, season, HAGG, AAGG, GP, ATP, HTP, HACG, AACG)
df_train <- df_modeling[which(df_modeling$season != 1819),]
df_test <- df_modeling[which(df_modeling$season == 1819),]
cat("Training set contains",nrow(df_train), "rows")
cat('\n')
cat("Training set contains",nrow(df_test), "rows")
```


##### Decision Tree

```{r}
set.seed(327)
epl_tree_cp <- train(
  FTR ~ .,
  data = df_train, 
  method = "rpart",
  trControl = trainControl(method = "cv",
                           number = 5),
  tuneGrid = data.frame(cp = 10^seq(-4, -2 , 
                                    length = 100)),
  na.action = na.omit
)
```

```{r}
epl_tree_cp$results
epl_tree_cp$results %>% 
  ggplot(aes(x = cp, y = Accuracy)) +
  geom_line()
```

*Discussion* The highest accuracy yielded by this decision tree was at 63.95% and it seems to level out at this point. This was reached as the complexity paramter increases towards .01.

```{r}
confusionMatrix(data = predict(epl_tree_cp, type = "raw"), #predictions
                reference = df_train$df_1819 , #actuals
                positive = "A") 
```


#### Bagging 

```{r}
set.seed(327)
epl_bagging <- train(
  FTR ~ .,
  data = df_train, 
  method = "rf",
  trControl = trainControl(method = "cv",
                           number = 5),
  tuneGrid = data.frame(mtry = 16),
  ntree = 100, #number of trees used, default is 500
  importance = TRUE, #for importance plots later
  nodesize = 5, #this is the default terminal node size for regression trees. Could set larger for smaller trees.
  na.action = na.omit
)
```

```{r}
epl_bagging$results
plot(epl_bagging$finalModel)
```
*Discussion* The accuracy for this bagging model comes out to be ~64.6% slightly higher than the decision tree. 

#### Random Forest
```{r}
set.seed(327)
epl_randf_oob <- train(
  FTR ~ .,
  data = df_train, 
  method = "rf",
  trControl = trainControl(method = "oob"),
  tuneGrid = data.frame(mtry = c(2,4,6,8,10,12)),
  ntree = 200, #number of trees used, default is 500
  importance = TRUE, #for importance plots later
  nodesize = 5, #this is the default terminal node size for regression trees. Could set larger for smaller trees.
  na.action = na.omit
)
```

```{r}
plot(epl_randf_oob$finalModel)
epl_randf_oob$bestTune
epl_randf_oob$results$Accuracy[3]
```

*Discussion* When we tune the random forest out-of-bag model to have the highest accuracy, we find that our model has an accuracy of 65.13%, this is just slightly higher than the previous bagging, and decision tree models. 

```{r}
ggplot(epl_randf_oob)
df_train
vip(epl_randf_oob$finalModel) 
```


#KNN

```{r}

set.seed(327)
epl_knn <- train(
  FTR ~., 
  data = df_train, 
  method = "knn", 
  trControl = trainControl(method = "cv",
                           number = 5,
                          # summaryFunction = model_stats,
                           returnResamp = "all"),
  tuneGrid = data.frame(k = c(1:15)))
  

epl_knn$results
epl_knn$bestTune$k
```

*Discussion* The best tuned KNN model (k = 6) yields an accurac of 62.6% lower than all three previous models. 

#### Lasso Model 

```{r}
set.seed(253)
all_vars_mod <- train(
    FTR ~ .,
    data = df_train,
    method = "glm",
    family = "binomial",
    trControl = trainControl(method = "cv", number = 5),
    metric = "Accuracy",
    na.action = na.omit
)

```

