---
title: "ml_epl"
author: "Alex Foster"
date: "4/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(ggplot2)
library(dplyr)
library(janitor)
library(reshape2)
library(tidyr)
library(readr)

library(rstan)
library(rstanarm)
library(bayesplot)
library(bayesrules)
library(lubridate)
library(Matrix)
```



#### Load in the data
```{R}
epl_1819 <- read_csv("season-1819_csv.csv") %>%
  mutate(season = 1819, Datetime = dmy(Date),HAGG = 0.0, AAGG=0.0, GP = 0, HTP = 0, ATP=0) %>%
  select(Datetime, HomeTeam, AwayTeam, FTR,FTAG, FTHG, HTAG, HTHG, HTR,season, HAGG, AAGG,GP, ATP, HTP)
epl_1718 <- read_csv("season-1718_csv.csv") %>%
  mutate(season = 1718, Datetime = dmy(Date),HAGG = 0.0, AAGG=0.0, GP = 0, HTP = 0, ATP=0) %>%
  select(Datetime, HomeTeam, AwayTeam, FTR,FTAG, FTHG, HTAG, HTHG, HTR,season, HAGG, AAGG,GP, ATP, HTP)
epl_1617 <- read_csv("season-1617_csv.csv") %>%
  mutate(season = 1617, Datetime = dmy(Date),HAGG = 0.0, AAGG=0.0, GP = 0, HTP = 0, ATP=0)%>%
  select(Datetime, HomeTeam, AwayTeam, FTR,FTAG, FTHG, HTAG, HTHG, HTR,season, HAGG, AAGG,GP, ATP, HTP)

seasons = rbind(epl_1617,epl_1718)
#seasons = rbind(epl_1718, epl_1819)
dim(seasons)
head(seasons)
```

#### Data Wrangling
* Need to make a dataframe where each row represents a game and has the following columns 
  * HT -- Home Team
  * AT -- Away Team
  * HTR -- Half Time Result
  * FTR -- Full Time Result
  * HTHG -- Half Time Home Goals
  * HTAG -- Half Time Away Goals
  * HAGG --  Home Average Goals per Game
  * AAGG -- Away Average Goals per Game 
  * HTP -- Home Team Position in the league
  * ATP -- Away Team Position in the league
  * Datetime -- lubricate datetime object 
* Response variable: FTR
* Explanatory variables: HTR, HTHG, HTAG, HAGG, AAGG, HTP, ATP 


```{r}
epl_1617$Datetime
seasons_df <- seasons

current_date <- seasons_df[1,]$Datetime
```

```{r}
# Function that computes the table of the season given a certain date in the season.
# Returns a dataframe that provides stats dependent on the progress of the given season 
compute_table <- function(date, df) {
  current_season = 1617
  df_subset <- df[which(df$season == current_season & df$Datetime <= date), ] # get games from that season and up to the given data
  
  # get team names
  names <- unique(df_subset$HomeTeam)
  
  # initialize beginning of season table
  table <- data.frame("Team" = names, "Points" = matrix(0,length(names)), "GA" = matrix(0,length(names)), "GF" = matrix(0,length(names)), 'GP' = matrix(0,length(names)))

  for (row in 1:nrow(df_subset)) {
    current_r =df_subset[row, ]
    # Update info for home team
    home <- current_r$HomeTeam
    table$GA[which(table$Team == home)] <- table$GA[which(table$Team == home)] + current_r$FTAG# GA
    table$GF[which(table$Team == home)] <- table$GF[which(table$Team == home)]+current_r$FTHG# GF
    table$GP[which(table$Team == home)] <- table$GP[which(table$Team == home)]+1

    # Update info for away team
    away <- current_r$AwayTeam
    table$GA[which(table$Team == away)] <- table$GA[which(table$Team == away)]+current_r$FTHG# GA
    table$GF[which(table$Team == away)] <- table$GF[which(table$Team == away)]+current_r$FTAG# GF
    table$GP[which(table$Team == away)] <- table$GP[which(table$Team == away)]+1

    # Award points to winner and for ties.
    if (current_r$FTR == 'H'){ # home wins
      table$Points[which(table$Team == home)] <- table$Points[which(table$Team == home)] + 3
    } else if (current_r$FTR == 'A'){ # away wins
      table$Points[which(table$Team == away)] <- table$Points[which(table$Team == away)] + 3
    } else { # draw
      table$Points[which(table$Team == home)] <- table$Points[which(table$Team == home)] + 1
      table$Points[which(table$Team == away)] <- table$Points[which(table$Team == away)] + 1
    }
  }
  # compute goal differential
  table <- table %>%
    mutate('GD' = GF - GA)
  # add AGSG column that represents the avg goals scored per game
  table <- table %>%
    mutate(AGSG = GF / GP)
  # add AGCG column that represents the avg goals conceded per game
  table <- table %>%
    mutate(AGCG = GA / GP)
  # Add position column that indicates what place each team is in.
  table <- table[with(table,order(table$Points, table$GD)),]
  table$Position <- length(names):1
  return(table)
}

```

  
```{r}
# Function takes in dataframe with all matches. Then loops through each match. For every match it computes the league standings prior to the given match. 
# Uses league standings to add more explanatory variables (HAGG and AAGG).
add_explantory_var <- function(df) {
    for (row in 1:nrow(df)){ # iterate through every game
        match = df[row,]
        current_date = match$Datetime
       # print(month(current_date))
        if (month(current_date) <  10 & month(current_date) >= 8){
            
          # compute table
            table <- compute_table(current_date, df)
           
          # get HAGG and AAGG for dateframe
            if (match$HomeTeam %in% table$Team) {
              #print(table$Position[which(table$Team == match$AwayTeam)])
              
              df[row,]$HAGG <- table$HGSG[which(table$Team == match$HomeTeam)]
             # df[row,]$HTP <- table$Position[which(table$Team == match$AwayTeam)]
            }
            if (match$AwayTeam %in% table$Team) {
              df[row,]$AAGG <- table$AGSG[which(table$Team == match$AwayTeam)]
             # df[row,]$ATP <- table$Position[which(table$Team == match$AwayTeam)]
            }
        } 
    }
 # print(row)
  return(df)
  
  }

seasons_df <- add_explantory_var(seasons_df)
head(seasons_df)
```

#### Data Visualizations

```{r}
# Distrbution of full time results 
seasons_df %>% 
  ggplot(aes(x = FTR)) +
  geom_bar() +
  theme_minimal()

# Distribution of full time results based on the HTR
seasons_df %>%
  ggplot(aes(x = FTR, fill = HTR)) +
  geom_bar() +
  scale_fill_hue() +
  theme_minimal()

# Examing goals scored by away and home teams by date and facet by season  
ggplot(seasons_df) +
 aes(x = Datetime, y = FTHG) +
 geom_line(size = 1L, colour = "#0c4c8a") +
 theme_minimal() +
 facet_wrap(vars(season))
ggplot(seasons_df) +
 aes(x = Datetime, y = FTAG) +
 geom_line(size = 1L, colour = "#0c4c8a") +
 theme_minimal() +
 facet_wrap(vars(season))

# Examine average goals scored as the season progresses
seasons_df %>%
  ggplot(aes(x=Datetime, y=HAGG)) +
  geom_line() +
  facet_wrap(~season)
seasons_df %>%
  ggplot(aes(x=Datetime, y=AAGG)) +
  geom_line() +
  facet_wrap(~season)


```
