---
title: "Final Project MLS"
author: "Alex Foster, Chris Bajek, Webster An"
date: "4/23/2020"
output: html_document
---

```{r setup, include=FALSE}

# Load packages
library(ggplot2)
library(dplyr)
library(janitor)
library(reshape2)
library(tidyr)
library(readr)
library(lubridate)
library(Matrix)
library(DataCombine)
library(rsample)
library(caret)
library(broom)

```
##### Analyze MLS Data with Machine Learning
#### Research Question: Can we predict the number of goals each team will score in a match?
* We intend on using multiple seasons. Our data source has data for the past 8 years. 
* Below we have begun adding explanatory variables and created visualizations for two seasons of matches.


#### Load in the data
```{R}
MLS <- read_csv("USA.csv")

# create datetime objects using lubridate. 
# also initialize variables for each game that will measure home/awayteam goals per game and home/away team goals conceded per game
MLS <- MLS %>%
  mutate(HAGG = 0.0, AAGG = 0.0, HACG = 0.0, AACG = 0.0, HTP=0, ATP = 0,TG = HG + AG, Datetime = dmy(Date),  month = month(Datetime)) %>%
  select(Country, League, Season, Datetime, Home, Away, HG, AG, Res, HAGG, AAGG, AACG, HACG,HTP,ATP,TG, PH, PD, PA)

ggplot(MLS, aes(Season)) + geom_bar() + ylab("Total Games") + ggtitle("MLS Dataset")

```

#### Data Wrangling
* We created a dataframe where each game in the original dataset is represented in two rows. One row for each team. 
  * Season -- MLS season that this match occured
  * Datetime -- lubridate datetime object
  * Team -- team that the row's statistics belong too
  * Opponent -- the other team in the match
  * Goals -- Number of goals the team scored
  * GoalsAllowed -- Number of goals the opponents scored on the team
  * Res -- result of the game, 'A'(away), 'H'(home) or 'D'(draw)
  * AGS -- Average Goals scored per match by the team in matches before this match in a given season 
  * Position -- Teams position in the league prior to this match
  * OpPosition -- Opponents position in the league prior to this match
  * TG -- Total Goals
  * HA -- 'H' or 'A' indicating if the team is the home team or away team
  * TableDiff -- Difference in positions between the two teams. 
  * OpAGC --  Average goals conceded per a match by opponents in matches prior to this one.
  * GoalScCo -- The sum of AGS and OpAGC
* Response variable: Goals
* Explanatory variables: Team, AGS, Position, OpPosition, HA, TableDiff, OpAGC, GoalScCo 


```{r, echo=FALSE}
# Function that computes the table of the season given a certain date in the season.
# Returns a dataframe that provides stats dependent on the progress of the given season 
compute_table <- function(date, df=USA) {
    # Get only the games that are in the same season and occured before the given date. 
    season = year(date)
    df_subset <- df[which(df$Season == season & df$Datetime < date), ] # get games from that season and up to the given data
    # get team names
    names <- unique(df[which(df$Season==season),]$Home)
    # initialize beginning of season table (stats are all zero)
    table <- data.frame("Team" = names, "Points" = matrix(0,length(names)), "GA" = matrix(0,length(names)), "GF" = matrix(0,length(names)), 'GP' = matrix(0,length(names)))
    if (nrow(df_subset) != 0){
        # Loop through each game and update the explanatory varaibles such as GA, GF, GP. All of these values are dependent on the date of the game. 
        for (row in 1:nrow(df_subset)) {
            current_r =df_subset[row, ]
            # Update info for home team
            home <- current_r$Home
            table$GA[which(table$Team == home)] <- table$GA[which(table$Team == home)] + current_r$AG# GA
            table$GF[which(table$Team == home)] <- table$GF[which(table$Team == home)]+current_r$HG# GF
            table$GP[which(table$Team == home)] <- table$GP[which(table$Team == home)]+1
            
            # Update info for away team
            away <- current_r$Away
            table$GA[which(table$Team == away)] <- table$GA[which(table$Team == away)]+current_r$HG# GA
            table$GF[which(table$Team == away)] <- table$GF[which(table$Team == away)]+current_r$AG# GF
            table$GP[which(table$Team == away)] <- table$GP[which(table$Team == away)]+1
            
            # Award points to winner and for ties. 
            if (current_r$Res == 'H'){ # home wins
              table$Points[which(table$Team == home)] <- table$Points[which(table$Team == home)] + 3
            } else if (current_r$Res == 'A'){ # away wins
              table$Points[which(table$Team == away)] <- table$Points[which(table$Team == away)] + 3
            } else { # draw
              table$Points[which(table$Team == home)] <- table$Points[which(table$Team == home)] + 1
              table$Points[which(table$Team == away)] <- table$Points[which(table$Team == away)] + 1
            }
        }
        # compute goal differential
        table <- table %>%
          mutate('GD' = GF - GA)
        # add AGSG column that represents the avg goals scored per game
        table <- table %>%
          mutate(AGSG = GF / GP)
        # add AGCG column that represents the avg goals conceded per game
        table <- table %>%
          mutate(AGCG = GA / GP)
        # Add position column that indicates what place each team is in. 
        table <- table[with(table,order(table$Points, table$GD)),]
        table$Position <- length(names):1
    }
  return(table)
}
```

```{r, echo=FALSE}
# Function computes table for every game and updates the current season stats like Avg goals scored, avg goals conceded and position in the league. 
add_table_stats <- function(df = usa_2012) {
    for (row in 1:nrow(df)){
      game = df[row,]
      # compute table 
      table <- compute_table(game$Datetime, df)
        # add home/away team average stats based on season performance up to that game.
        if (table[1,]$GP != 1){ # if games played is 1 then we don't really want average stats (avoids potential errors)
            if(game$Home %in% table$Team & game$Away %in% table$Team){ # check that both teams are actually in table.
                # home team stats
                df[row,]$HAGG <- table$AGSG[which(table$Team == game$Home)]
                df[row,]$HACG <- table$AGCG[which(table$Team == game$Home)]
                df[row,]$HTP <- table$Position[which(table$Team == game$Home)]
                # away teams stats
                df[row,]$AAGG <- table$AGSG[which(table$Team == game$Away)]
                df[row,]$AACG <- table$AGCG[which(table$Team == game$Away)]
                df[row,]$ATP <- table$Position[which(table$Team == game$Away)]
            }
        }
    }
    return(df)
}
```
 
```{r,echo=FALSE}
# initialize final dataframe with 2012 season
USA_final <- add_table_stats(MLS[which(MLS$Season==2012),])
# loop through all other seasons that aren't 2012 and compute season statistics.
for (season in unique(MLS$Season)){
  print(season)
  if (season != 2012){
    updated_df <- add_table_stats(MLS[which(MLS$Season==season),])
    USA_final <- rbind(USA_final, updated_df)
  }
}

write.csv(USA_final,"USA_final.csv", row.names = TRUE)
```

```{r,echo=FALSE, message=FALSE}
# Read in dataset (avoid long code run)
MLS_df <- read_csv("USA_final.csv",)
MLS_df <- MLS_df[which(MLS_df$PH != 'USA'),]
MLS_df <- subset(MLS_df,select= -c(X1))
```
##### Putting it all together
```{r}
Home <- MLS_df
Home$Country <- NULL
Home$League <- NULL
Home$Date <- NULL
Home$Time <- NULL
# Home$Away <- NULL
Home$AAGG <- NULL
Home$HACG <- NULL
Home <- Home %>% 
  rename(
    Team = Home,
    Opponent = Away,
    Goals = HG,
    GoalsAllowed = AG,
    Position = HTP,
    AGS = HAGG,
    OpAGC = AACG,
    OpPosition = ATP
    )
Home <- cbind(Home, HA = "H")
Away <- MLS_df
Away$Country <- NULL
Away$League <- NULL
Away$Date <- NULL
Away$Time <- NULL
# Away$Home <- NULL
Away$HAGG <- NULL
Away$AACG <- NULL
Away <- Away %>% 
  rename(
    Team = Away,
    Opponent = Home,
    Goals = AG,
    GoalsAllowed = HG,
    Position = ATP,
    AGS = AAGG,
    OpAGC = HACG,
    OpPosition = HTP,
    Opponent = Home
    )
Away <- cbind(Away, HA = "A")
MLS_final <- rbind(Home, Away)

MLS_final[,"Position"] <- sapply(MLS_final[, "Position"], as.numeric)
MLS_final[,"OpPosition"] <- sapply(MLS_final[, "OpPosition"], as.numeric)
MLS_final[,"AGS"] <- sapply(MLS_final[, "AGS"], as.numeric)
MLS_final[,"OpAGC"] <- sapply(MLS_final[, "OpAGC"], as.numeric)
MLS_final <- MLS_final %>%
  mutate(TableDiff = Position - OpPosition,
         GoalDiff = Goals - GoalsAllowed,
         GoalScCo = AGS + OpAGC)
if ('X1' %in% names(MLS_final)){
  MLS_df[ , !(names(MLS_df) %in% c('X1'))]

}

#Accounting for the fact that Atlanta United has two different names "Atlanta United" and "Atlanta Utd"

# Create replacements data frame
Replaces <- data.frame(from = "Atlanta Utd", to = "Atlanta United")

# Replace patterns and return full data frame
MLS_final1 <- FindReplace(data = MLS_final, Var = "Team", replaceData = Replaces,
                     from = "from", to = "to", exact = FALSE)
MLS_final <- MLS_final1
remove(MLS_final1)
head(MLS_final)
```

#### Data Visualizations

```{r}
# Distrbution of goals by season (broken into months)
MLS_final %>% 
  ggplot(aes(x = Goals)) +
  geom_bar(colour = "#0c4c8a") +
  facet_wrap(~Season)+
  theme_minimal() +
  ggtitle("Goals by Season (Broken in to Months)") +
  theme(plot.title = element_text(hjust = 0.5))

# Distribution of full time results  by season
MLS_final %>%
  ggplot(aes(x = Res)) +
  geom_bar() +
  scale_fill_hue() +
  facet_wrap(~Season)+
  theme_minimal() +
  ggtitle("Full Time Results by Season") +
  theme(plot.title = element_text(hjust = 0.5))

# Total Goals by team
MLS_final %>%
  ggplot(aes(x = Team, y = Goals)) +
  geom_col() +
  scale_fill_hue() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ggtitle("Total Goals by Team") +
  theme(plot.title = element_text(hjust = 0.5))

# Examining average goals scored
MLS_final %>%
  group_by(Team) %>% 
  mutate(avg_goals=AGS/n()) %>% 
  ggplot(aes(x = Team, y = avg_goals)) +
  geom_col() +
  scale_fill_hue() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ylab("Average Goals per Game") +
  ggtitle("Average Goals by Team") +
  theme(plot.title = element_text(hjust = 0.5))

# Examining relationship between AGS and Position by team
MLS_final %>% 
  ggplot(aes(x=AGS, y=Position)) +
  geom_point(alpha=.5) +
  scale_fill_hue() +
  facet_wrap(~Team)+
  theme_minimal() +
  ggtitle("Average Goals by Table Position") +
  theme(plot.title = element_text(hjust = 0.5))

# Number of goals scored by team during home games or away games
MLS_final %>% 
  ggplot(aes(x=HA, y=Goals)) +
  geom_col() +
  scale_fill_hue() +
  facet_wrap(~Team)+
  theme_minimal() +
  ggtitle("Total Goals Scored at Home and Away") +
  theme(plot.title = element_text(hjust = 0.5))
```

#### Modeling 
 

Predict Goals Variable

##### Create training/testing set
* Going to have most recent season be testing set.
```{r}
df_modeling <- MLS_final
df_train <- df_modeling[which(df_modeling$Season != 2019),]
df_train <- df_modeling[which(df_train$Season != 2020),] # remove 2020 from training set.
df_test <- df_modeling[which(df_modeling$Season == 2019),]
cat("Training set contains",nrow(df_train), "rows")
cat('\n')
cat("Training set contains",nrow(df_test), "rows")
```


##### OLS
```{r}
set.seed(123) 
split <- trainControl(method = "cv", number = 5)

ols_model <- train(
  Goals ~ Team + AGS + Position + OpPosition + HA + OpAGC,
  data = df_train,
  method = "lm", 
  trControl = split,
  na.action = na.omit
)
summary(ols_model)
```



##### Stepwise Selection of Variables
```{r}
set.seed(123) 
split <- trainControl(method = "cv", number = 5)

cv_cars_all_vars <- train(
  MPG ~ .,
  data = cars_train %>% select(-`Model Index`, -Model),
  method = "leapForward", 
  tuneGrid = data.frame(nvmax = 1:16), 
  trControl = split,
  na.action = na.omit
)
```



##### Lasso Model 
```{r}
set.seed(253)
all_vars_mod <- train(
    Goals ~ Team + AGS + Position + OpPosition + HA + TableDiff + OpAGC + GoalScCo,
    data = MLS_final,
    method = "glm",
    family = "binomial",
    trControl = trainControl(method = "cv", number = 5),
    metric = "Accuracy",
    na.action = na.omit
)

```


##### Random Forest
```{r}
set.seed(327)
epl_randf_oob <- train(
  FTR ~ .,
  data = df_train, 
  method = "rf",
  trControl = trainControl(method = "oob"),
  tuneGrid = data.frame(mtry = c(2,4,6,8,10,12)),
  ntree = 200, #number of trees used, default is 500
  importance = TRUE, #for importance plots later
  nodesize = 5, #this is the default terminal node size for regression trees. Could set larger for smaller trees.
  na.action = na.omit
)
```

```{r}
plot(epl_randf_oob$finalModel)
epl_randf_oob$bestTune
epl_randf_oob$results$Accuracy[3]
```

*Discussion* When we tune the random forest out-of-bag model to have the highest accuracy, we find that our model has an accuracy of 65.13%, this is just slightly higher than the previous bagging, and decision tree models. 

```{r}
ggplot(epl_randf_oob)
df_train
vip(epl_randf_oob$finalModel) 
```


##### KNN

```{r}

set.seed(327)
epl_knn <- train(
  FTR ~., 
  data = df_train, 
  method = "knn", 
  trControl = trainControl(method = "cv",
                           number = 5,
                          # summaryFunction = model_stats,
                           returnResamp = "all"),
  tuneGrid = data.frame(k = c(1:15)))
  

epl_knn$results
epl_knn$bestTune$k
```

*Discussion* The best tuned KNN model (k = 6) yields an accurac of 62.6% lower than all three previous models. 


